---
title: "CLUSTERING"
author: "Cindy Njeri Kiarie"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Clustering and Optimization Techniques"
output: html_document
---

# Question 1: Clustering Methods

## 1. EM (Expectation-Maximization) Clustering

The EM algorithm is widely used for clustering tasks where data is modeled as coming from a mixture of distributions. It optimizes the likelihood of the observed data by iteratively refining the parameters of these distributions.

### Overview

- Assumes the data points are generated from a mixture of probability distributions.
- Commonly uses Gaussian distributions for the clusters.

### Steps in EM Clustering

1. **Initialization**  
   - Define the number of clusters (\(k\)).  
   - Initialize the parameters of each distribution, such as means (\(\mu\)), covariances (\(\Sigma\)), and mixing coefficients (\(\pi\)).

2. **Expectation (E) Step**  
   - Compute the likelihood of each data point belonging to a specific cluster. This is done using Bayes’ theorem based on the current parameters.

3. **Maximization (M) Step**  
   - Update the parameters (\(\mu\), \(\Sigma\), \(\pi\)) to maximize the overall likelihood of the data.

4. Repeat E and M steps until convergence or until the likelihood does not significantly improve.

### Applications

- Image segmentation  
- Anomaly detection  
- Customer segmentation

### Example Dataset

Use the **Iris dataset** (4 features: sepal length, sepal width, petal length, petal width) for Gaussian clustering.

```{r em-gaussian-clustering-iris, message=FALSE, warning=FALSE}
# Load required library
library(mclust)

# Load the Iris dataset
data(iris)
iris_features <- iris[, 1:4]  # Extract the four numeric features

# Fit a Gaussian Mixture Model using the EM algorithm
gmm_fit <- Mclust(iris_features)

# Display a summary of the fitted model
summary(gmm_fit)

# Add the predicted cluster labels to the original data
iris$Cluster <- as.factor(gmm_fit$classification)

# Plot the classification result
plot(gmm_fit, what = "classification")

# View the first few rows of the clustered data
head(iris)
```

---

## 2. K-Means / K-Median Clustering

K-Means and K-Median are partition-based clustering methods focusing on minimizing the distance between data points and their cluster centers.

### K-Means Clustering

- **Objective Function**: Minimize the sum of squared Euclidean distances between data points and their cluster centroids.

#### Steps

1. Initialize \(k\) cluster centroids randomly.  
2. Assign each data point to the nearest centroid.  
3. Update each centroid to the mean of its assigned points.  
4. Repeat steps 2 and 3 until centroids stabilize.

### K-Median Clustering

- Similar to K-Means but minimizes the sum of the absolute differences (Manhattan distance).
- Uses **median** instead of **mean** to update cluster centers, making it robust to outliers.

### Applications

- Market segmentation  
- Document classification  
- Recommendation systems

### Example Dataset

```{r kmeans-kmedian-usarrests, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(ClusterR)  # For K-Median clustering

# Load the USArrests dataset
data("USArrests")
usarrests_data <- na.omit(USArrests)  # Just in case

# --- K-MEANS CLUSTERING ---
set.seed(123)
kmeans_result <- kmeans(usarrests_data, centers = 3)

# Add K-Means cluster labels
usarrests_kmeans <- usarrests_data %>%
  mutate(Cluster = as.factor(kmeans_result$cluster))

# Plot K-Means clusters (first two principal components)
usarrests_pca <- prcomp(usarrests_data, scale. = TRUE)
pca_df <- as.data.frame(usarrests_pca$x)
pca_df$Cluster <- usarrests_kmeans$Cluster

ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3) +
  labs(title = "K-Means Clustering on USArrests", x = "PC1", y = "PC2") +
  theme_minimal()

# --- K-MEDIAN CLUSTERING ---
set.seed(123)
# Create a matrix for initial centroids
initial_centroids <- matrix(data = c(8, 170, 65, 20, 
                                     5, 120, 70, 18, 
                                     3, 100, 80, 16), 
                            nrow = 3, ncol = 4, byrow = TRUE)

# Pass this matrix to the CENTROIDS argument
kmed_result <- KMeans_arma(data = scale(usarrests_data), clusters = 3, 
                           n_iter = 100, CENTROIDS = initial_centroids, 
                           seed_mode = "keep_existing")

# Calculate Euclidean distances from each point to each centroid
distances <- as.matrix(dist(scale(usarrests_data)))  # Scale the data to match centroids
cluster_assignments <- apply(scale(usarrests_data), 1, function(row) {
  which.min(colSums((t(kmed_result) - row)^2))
})

# Add the cluster labels to the dataset
usarrests_kmed <- usarrests_data %>%
  mutate(Cluster = as.factor(cluster_assignments))

# Add K-Median cluster labels
usarrests_kmed <- usarrests_data %>%
  mutate(Cluster = as.factor(cluster_assignments))

# Add the cluster assignments to the PCA data frame
pca_df$Cluster <- as.factor(cluster_assignments)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3) +
  labs(title = "K-Median Clustering on USArrests", x = "PC1", y = "PC2") +
  theme_minimal()
```


---

## 3. DBSCAN (Density-Based Spatial Clustering of Applications with Noise)

DBSCAN is a density-based method that identifies clusters based on areas of high data point density, making it effective for non-spherical clusters and noisy data.

### Key Parameters

1. \( \varepsilon \): Radius defining the neighborhood of a data point.  
2. **minPts**: Minimum number of points required to form a dense region (cluster).

### Steps

1. For each data point, calculate its neighborhood (points within radius \( \varepsilon \)).  
2. Classify data points as:  
   - **Core Point**: Has at least \( \text{minPts} \) neighbors.  
   - **Border Point**: Lies in the neighborhood of a core point but has fewer than \( \text{minPts} \) neighbors.  
   - **Noise Point**: Neither core nor border point.  

3. Expand clusters from core points, including all reachable core and border points.  
4. Label noise points as outliers.

### Applications

- Geospatial data analysis  
- Fraud detection  
- Clustering irregular shapes in image data

### Example Dataset

```{r dbscan-iris, message=FALSE, warning=FALSE}
# Load required libraries
library(dbscan)
library(tidyverse)
library(factoextra)  # For PCA and visualization

# Load the iris dataset
data("iris")
iris_features <- iris[, 1:4]

# Scale the data
iris_scaled <- scale(iris_features)

# Estimate eps using k-NN distance plot (optional visual aid)
kNNdistplot(iris_scaled, k = 5)
abline(h = 0.6, lty = 2, col = "red")

# Apply DBSCAN
dbscan_result <- dbscan(iris_scaled, eps = 0.6, minPts = 5)

# Add cluster labels to the dataset
iris$DBSCAN_Cluster <- as.factor(dbscan_result$cluster)

# Reduce to 2D using PCA for plotting
pca_res <- prcomp(iris_scaled)
pca_df <- as.data.frame(pca_res$x[, 1:2])
pca_df$Cluster <- iris$DBSCAN_Cluster

# Plot DBSCAN clustering result
ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3) +
  labs(title = "DBSCAN Clustering on Iris Dataset", x = "PC1", y = "PC2") +
  theme_minimal()
```


---

# Question 2: Other Clustering Methods

## Agglomerative Clustering

- Builds a hierarchy of clusters by starting with each data point as its own cluster and successively merging the closest pairs of clusters.
- Merging is based on a similarity metric (e.g., single linkage, complete linkage, or average linkage).
- Output can be visualized using a **dendrogram**.

## Mean Shift Clustering

- Iteratively shifts data points towards the densest area in the feature space (density peaks).
- The number of clusters is determined automatically based on the data distribution.

## Spectral Clustering

- Constructs a similarity graph where nodes represent data points and edges are weighted by a similarity measure.
- Applies **eigenvalue decomposition** to this graph and performs clustering (e.g., K-Means) in the transformed space.

---

# Question 3: Gradient Operator and Gradient Descent

## Gradient Operator (\( \nabla \))

The gradient operator is a vector of partial derivatives that points in the direction of the steepest ascent of a scalar function.

For a function \( f(x_1, x_2, \dots, x_n) \), the gradient is:

\[
\nabla f = \left( \frac{\partial f}{\partial x_1}, \frac{\partial f}{\partial x_2}, \dots, \frac{\partial f}{\partial x_n} \right)
\]

### Properties

- Gradient magnitude represents the rate of change.  
- Direction of \( \nabla f \) is perpendicular to the level sets of \( f \).

---

## Gradient Descent

Gradient Descent is an iterative optimization algorithm for minimizing a function.

### Steps

1. Initialize parameters (\( \theta \)) randomly.  
2. Compute the gradient (\( \nabla f(\theta) \)).  
3. Update parameters:  

\[
\theta = \theta - \eta \nabla f(\theta)
\]

where \( \eta \) is the learning rate.  

4. Repeat steps 2 and 3 until convergence.

### Variants

1. **Batch Gradient Descent**: Computes gradient using the entire dataset.  
2. **Stochastic Gradient Descent (SGD)**: Computes gradient using one sample at a time.  
3. **Mini-batch Gradient Descent**: Uses a small subset of the data for computing gradients.

### Applications

- Training machine learning models  
- Optimizing cost functions
